---
title: 'Project #'
author: "Santiago Castro Dau, June Monge, Rachita Kumar, Sarah Lötscher"
output: pdf_document
---

## Problem 23: d-separation 
###  1. Write down all the variables that are d-separated from A given {C,D}
A and G are d-separated since both the paths A-C-G and A-D-G are both blocked 

###  2. Indicate whether each statement is true or false and explain your choice.
#### (a) B is conditionally independent of C given D.
False--> the path B-D-A-C is still active
#### (b) G is conditionally independent of E given D.
False--> the path G-C-A-D-B-F-E is still active
#### (c) C is conditionally independent of F given A.
True--> true since all paths are inactive
#### (d) C is conditionally independent of E given its Markov blanket (of C)
True --> Generally for any node if  conditioned on the markov blanket its going to be independend of the other nodes. Every node is independent of its non-descendants(parents).


## Problem 24: Testing for marginal correlation 
###  Using the data from MVN DAG.rds2, display the observations of A and B in a scatterplot. 
```{r}
library("ggplot2")
```

```{r}
#Read in dataset
set.seed(1)
DAG<-readRDS("./MVN_DAG.rds", refhook = NULL)

```

```{r}
#Create the scatterplot

slr.coef<-coef(lm(B~ A,DAG))
qplot(A,B,data=DAG)+theme_classic()+geom_abline(intercept =slr.coef[1], slope = slr.coef[2],colour="blue" )+ ggtitle("Marginal coorelation")


```
What does the plot suggest about their (marginal) correlation? Does it agree with Figure 3?
TBA........

```{r}
#Coorelation Test

cor.test(DAG$A,DAG$B)

```

Use the function cor.test() to test the null hypothesis of no correlation between A and B. What is your conclusion?
We cant reject the hypothesis that A and B are uncorrelated


