---
title: "Project 3"
author: "Santiago Castro Dau, June Monge, Rachita Kumar, Sarah Lötscher"
date: "3/14/2022"
output: pdf_document
---

## Problem 8:  Profile HMMs: Estimating match emission probabilities
### What are the estimated match emission probabilities of the profile HMM in Figure 1?

The model
```{r}
## The Example Data

bat <- c("A","G","-","-","-","C")
rat <- c("A","-","A","G","-","C")
cat <- c("A","G","-","-","-","C")
gnat <- c("-","G","A","A","A","C")
goat <- c("A","G","-","-","A","C")
M <- rbind(bat,rat,cat,gnat,goat)

# Alphabet
A=c("A","C","G","T")

```


Find out which positions are in Match state and which are in Insert state

```{r}
# Get the positions which are Insertstate and Match state
match = which(colSums(M!="-")>(dim(M)[1]/2)) #Vector with Match positions
insertion = c(1:dim(M)[2])[!c(1:dim(M)[2]) %in% match] #Vector with insert positions

```


```{r}
#Matrix of all pos????????????????
E=matrix(,0,dim(M)[2])
for (n in A){
  e<-apply(M==n,2,sum)
  E <- rbind(E,e)
}
rownames(E)=A


```

Calculate the Emission probabilities of a match

```{r}
# Emission probabilities of match
E_match_prob=t(t(E[,match]+1)/colSums(E[,match]+1))
E_match_prob

```



## Problem 9: Estimating insert emission probabilities
What are the estimated insert emission probabilities of the profile HMM in Figure 1?

```{r}

consecutive_in=split(insertion, cumsum(c(1, diff(insertion) != 1)))
E_cons=matrix(,length(A),0)
for (insert in consecutive_in){
  s = rowSums(E[,insert])
  E_cons = cbind(E_cons,s)
}
rownames(E_cons)=A

# Emission probabilities of insertion
E_insertion_prob=t(t(E_cons+1)/colSums(E_cons+1))
E_insertion_prob

```


## Problem 10: Estimating transition probabilities
What are the estimated transmission probabilities in the profile HMM in Figure 1?

```{r}
#Transition probabilities
counts.only=FALSE

#Get state Matrix
SM=matrix(FALSE,dim(M)[1],dim(M)[2])
SM[,match][M[,match]!='-']='M'
SM[,match][M[,match]=='-']='D'
SM[,insertion][M[,insertion]!='-']='I'

#Initate Transition Matrix
transitions <- c("MM","MD","MI","IM","ID","II","DM","DD","DI")
Tr=matrix(0,length(transitions),dim(M)[2]+1)
rownames(Tr)=transitions


#Create the Transition counts for every possible szenario
for (i in 1:dim(SM)[1]){
  prev <- 'M' 	
  prevStateNum <- 0 
  for(j in 1:(dim(SM)[2])) {
    newState<- SM[i,j]
    if (newState==FALSE){
      prevStateNum <- j
      next
    }
    transition <- paste(prev, newState, sep="")
    Tr[transition, prevStateNum +1] <- Tr[transition, prevStateNum +1] +1
    prevStateNum <- j
    prev <- newState
  }
  transition<- paste(prev, 'M', sep="")
  Tr[transition, prevStateNum +1] <- Tr[transition, prevStateNum +1] +1
}

#Add the Insertion state columns to the corresponding match state
for (insert in rev(consecutive_in)){
  s = rowSums(Tr[,c(insert[1]-1,insert)])
  Tr[,insert[1]-1]=s
  Tr=Tr[,-insert] #Delete the insertion columns
}


#Add pseudo-count and Get the Transition probabilities
if (!counts.only) {
    Tr <- Tr+1
    for(i in 1:ncol(Tr)) {
      Tr[1:3,i] <- Tr[1:3,i] / sum(Tr[1:3,i])
      Tr[4:6,i] <- Tr[4:6,i] / sum(Tr[4:6,i])
      Tr[7:9,i] <- Tr[7:9,i] / sum(Tr[7:9,i])
    }
}

colnames(Tr)=c(0:(ncol(Tr)-1))
Tr

```


## Problem 11: Protein family membership classification


## (1) Import functions

```{r}
#Get Functions
source("./code/profileHMM.R")
```

## (2) Read the two alignments

```{r}
#Read in Data
GTP_binding_proteins = parseAlignment("./data/GTP_binding_proteins.txt")
ATPases = parseAlignment("./data/ATPases.txt")

```
## (3) Parametrise two profile HMMs

Parametrise two profile HMMs for each protein family

```{r}
#Get the profile HMMS

profileHMM_GTP = learnHMM(GTP_binding_proteins)
profileHMM_ATPases = learnHMM(ATPases)

```


## (4) Highest match and highest insert emission frequencies for ATPases


```{r,tidy=TRUE}
#Highest match emission frequencies
match_frequency_positions_ATPases = 
  which(profileHMM_ATPases$mE == max(profileHMM_ATPases$mE,na.rm=TRUE), 
        arr.ind = TRUE)[2]
print(match_frequency_positions_ATPases)

barplot(profileHMM_ATPases$mE[,match_frequency_positions_ATPases],
        names.arg=rownames(profileHMM_ATPases), 
        main = paste0("Highest match emission frequency for ATPases at position: ",
                      match_frequency_positions_ATPases))

#Highest insert emission frequencies
insert_frequency_positions_ATPases = 
  which(profileHMM_ATPases$iE == max(profileHMM_ATPases$iE,na.rm=TRUE), 
        arr.ind = TRUE)[2]
print(insert_frequency_positions_ATPases)

barplot(profileHMM_ATPases$mE[,insert_frequency_positions_ATPases],
        names.arg=rownames(profileHMM_ATPases), 
        main = paste0("Highest insert emission frequency for ATPases at position: ",
                      insert_frequency_positions_ATPases))

```


## (4) Highest match and highest insert emission frequencies for GTP


```{r, tidy=TRUE}
#Highest match emission frequencies
match_frequency_positions_GTP = 
  which(profileHMM_GTP$mE == max(profileHMM_GTP$mE,na.rm=TRUE), 
                                      arr.ind = TRUE)[2]
print(match_frequency_positions_GTP)

barplot(profileHMM_GTP$mE[,match_frequency_positions_GTP],
        names.arg=rownames(profileHMM_GTP), 
        main = paste0("Highest match emission frequency for GTP at position: ",
                      match_frequency_positions_GTP))

#Highest insert emission frequencies
insert_frequency_positions_GTP = 
  which(profileHMM_GTP$iE == max(profileHMM_GTP$iE,na.rm=TRUE), 
                                       arr.ind = TRUE)[2]
print(insert_frequency_positions_GTP)

barplot(profileHMM_GTP$mE[,insert_frequency_positions_GTP],
        names.arg=rownames(profileHMM_GTP), 
        main = paste0("Highest insert emission frequency for GTP at position: ",
                      insert_frequency_positions_GTP))
```


## (5) Load the protein sequences into a list 

```{r}
#Load the protein sequence

unclassifiedProteins <- parseProteins(proteinsFile = 
                                        "./data/Unclassified_proteins.txt" )

```

## (6) Obtain the log odds ratio and plot the results


```{r, tidy=TRUE}
P_ATPases <- lapply(X = unclassifiedProteins, forward, HMM=profileHMM_ATPases)
P_GTP <- lapply(X = unclassifiedProteins, forward, HMM=profileHMM_GTP)
#Log-Odds Ratio of ATPases Profile HMM wrt GTP Profile HMM
q <- unlist(P_ATPases)-unlist(P_GTP)
plot(q,col=ifelse(q>=0,"red","blue"), 
     main="Log Odds Ratio of ATPases Profile HMM wrt GTP Profile HMM",
     ylab = "Log Odds Ratio", xlab = "Sequence Index")
```


```{r, tidy=TRUE}
unclassifiedATPases = as.numeric(which(q >=0,arr.ind = TRUE))
no_unclassifiedATPases = length(unclassifiedATPases)
print(paste0("The number of predicted ATPases is ",no_unclassifiedATPases))
print("The unclassified proteins that are ATPases have indices")
print(unclassifiedATPases)

unclassifiedGTP = as.numeric(which(q < 0,arr.ind = TRUE))
no_unclassifiedGTP = length(unclassifiedGTP)
print(paste0("The number of predicted GTPs is ",no_unclassifiedGTP))
print("The unclassified proteins that are GTPs have indices")
print(unclassifiedGTP)
```


```{r}
# #TO DO
# library(ggplot2)
# 
# logsodd<-vector()
# 
# #loop through all the sequences in protdd
# for(i in 1:length(Protdd)){
#   loGTP<-forward(GTPHMM,unlist(Protdd[i]))
#   loATP<-forward(ATPHMM,unlist(Protdd[i]))
#   logsodd <- c(logsodd,loGTP[[1]]-loATP[[1]])
#   
# }
# logsodd
# 
# 
# a <- barplot(logsodd,
#             col= ifelse(logsodd < 0,"red","blue"),
#             xaxt = "n", yaxt = "n"
# )
# 
# 
# y<-vector(1:length(logsodd))
# 
# ggplot(logsodd) +geom_histogram()

```



