---
title: "Project 3"
author: "Santiago Castro Dau, June Monge, Rachita Kumar, Sarah Lötscher"
date: "3/14/2022"
output: pdf_document
---

## Problem 8:  Profile HMMs: Estimating match emission probabilities
### What are the estimated match emission probabilities of the profile HMM in Figure 1?

The model
```{r}
## The Example Data

bat <- c("A","G","-","-","-","C")
rat <- c("A","-","A","G","-","C")
cat <- c("A","G","-","-","-","C")
gnat <- c("-","G","A","A","A","C")
goat <- c("A","G","-","-","A","C")
M <- rbind(bat,rat,cat,gnat,goat)

# Alphabet
A=c("A","C","G","T")

```


Find out which positions are in Match state and which are in Insert state

```{r}
# Get the positions which are Insertstate and Match state
match = which(colSums(M!="-")>(dim(M)[1]/2)) #Vector with Match positions
insertion = c(1:dim(M)[2])[!c(1:dim(M)[2]) %in% match] #Vector with insert positions

```


```{r}
#Matrix of all pos????????????????
E=matrix(,0,dim(M)[2])
for (n in A){
  e<-apply(M==n,2,sum)
  E <- rbind(E,e)
}
rownames(E)=A


```

Calculate the Emission probabilities of a match

```{r}
# Emission probabilities of match
E_match_prob=t(t(E[,match]+1)/colSums(E[,match]+1))
E_match_prob

```



## Problem 9: Estimating insert emission probabilities
What are the estimated insert emission probabilities of the profile HMM in Figure 1?

```{r}

consecutive_in=split(insertion, cumsum(c(1, diff(insertion) != 1)))
E_cons=matrix(,length(A),0)
for (insert in consecutive_in){
  s = rowSums(E[,insert])
  E_cons = cbind(E_cons,s)
}
rownames(E_cons)=A

# Emission probabilities of insertion
E_insertion_prob=t(t(E_cons+1)/colSums(E_cons+1))
E_insertion_prob

```


## Problem 10: Estimating transition probabilities
What are the estimated transmission probabilities in the profile HMM in Figure 1?

```{r}
#Transition probabilities
counts.only=FALSE

#Get state Matrix
SM=matrix(FALSE,dim(M)[1],dim(M)[2])
SM[,match][M[,match]!='-']='M'
SM[,match][M[,match]=='-']='D'
SM[,insertion][M[,insertion]!='-']='I'

#Initate Transition Matrix
transitions <- c("MM","MD","MI","IM","ID","II","DM","DD","DI")
Tr=matrix(0,length(transitions),dim(M)[2]+1)
rownames(Tr)=transitions


#Create the Transition counts for every possible szenario
for (i in 1:dim(SM)[1]){
  prev <- 'M' 	
  prevStateNum <- 0 
  for(j in 1:(dim(SM)[2])) {
    newState<- SM[i,j]
    if (newState==FALSE){
      prevStateNum <- j
      next
    }
    transition <- paste(prev, newState, sep="")
    Tr[transition, prevStateNum +1] <- Tr[transition, prevStateNum +1] +1
    prevStateNum <- j
    prev <- newState
  }
  transition<- paste(prev, 'M', sep="")
  Tr[transition, prevStateNum +1] <- Tr[transition, prevStateNum +1] +1
}

#Add the Insertion state columns to the corresponding match state
for (insert in rev(consecutive_in)){
  s = rowSums(Tr[,c(insert[1]-1,insert)])
  Tr[,insert[1]-1]=s
  Tr=Tr[,-insert] #Delete the insertion columns
}


#Add pseudo-count and Get the Transition probabilities
if (!counts.only) {
    Tr <- Tr+1
    for(i in 1:ncol(Tr)) {
      Tr[1:3,i] <- Tr[1:3,i] / sum(Tr[1:3,i])
      Tr[4:6,i] <- Tr[4:6,i] / sum(Tr[4:6,i])
      Tr[7:9,i] <- Tr[7:9,i] / sum(Tr[7:9,i])
    }
}

Tr

```


## Problem 11: Protein family membership classification


## (1) Import functions

```{r}
#Get Functions
source("./code/profileHMM.R")
```

## (2) Read the two alignments

```{r}
#Read in Data
GTPdd<-parseAlignment("./data/GTP_binding_proteins.txt")

ATPdd<-parseAlignment("./data/ATPases.txt")

```
## (3) Parametrise two profile HMMs

Parametrise two profile HMMs for each protein family

```{r}
#Get the profile HMMS

GTPHMM<-learnHMM(GTPdd)

ATPHMM<-learnHMM(GTPdd)

```


## (4) Highest match and highest insert emission frequencies


```{r}
#TO DO


```

## (5) Load the protein sequences into a list 

```{r}
#Load the protein sequence

Protdd<-parseProteins("./data/Unclassified_proteins.txt")

```

## (6) Obtain the log odds ratio and plot the results


```{r}
#TO DO
library(ggplot2)

logsodd<-vector()

#loop through all the sequences in protdd
for(i in 1:length(Protdd)){
  loGTP<-forward(GTPHMM,unlist(Protdd[i]))
  loATP<-forward(ATPHMM,unlist(Protdd[i]))
  logsodd <- c(logsodd,loGTP[[1]]-loATP[[1]])
  
}
logsodd


a <- barplot(logsodd,
            col= ifelse(logsodd < 0,"red","blue"),
            xaxt = "n", yaxt = "n"
)


y<-vector(1:length(logsodd))

ggplot(logsodd) +geom_histogram()

```



