---
title: "Project 1. Solution"
author: "Santiago Castro Dau, June Monge, Rachita Kumar, Sarus"
date: "2/25/2022"
output: pdf_document
---
#HI
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(vroom)
library(BiDAG)
library(graph)
library(parallel)
```

## Problem 1: Conditional independence and BNs
\subsection*{a)}
The condition $A \perp B|C$ holds for this Bayesian network a).
Proof:
\begin{equation*}
\begin{split}
    P(A,B,C) &= P(A|C)P(B|C)P(C)\\
    \textnormal{Consider}\\
    P(A,B|C) &= \frac{P(A,B,C)}{P(C)}\\
    &=P(A|C)P(B|C)\\
    &\implies A\perp B|C\\
\end{split} 
\end{equation*}
Hence Proved.
\subsection*{b)}
The condition $A \perp B$ holds for this Bayesian network b).
Proof:
\begin{equation*}
\begin{split}
    P(A,B,C) &= P(A)P(B)P(C|A,B)\\
    P(C|A,B) &= \frac{P(A,B,C)}{P(A,B)}\\
    \textnormal{Substituting}\\
    P(A,B) &= P(A)P(B)\\    
    &\implies A\perp B\\
\end{split} 
\end{equation*}
Hence Proved.

## Problem 2: Markov blanket

## Problem 3: Learning Bayesian networks from protein data
### a)
```{r}
## Set seed
set.seed(2022)

## Read and index data
data <- vroom(file = "sachs.data.txt", show_col_types = FALSE)
data = data %>% rowid_to_column(var = 'row_id')

## Visualize data dimensions. 
dim(data)
```

* $N = 853$
* $n = 11$

```{r}
## Split data
train = data %>% sample_frac(0.8, replace = FALSE) 
test = anti_join(data, train, by = "row_id")

## Remove index column used for splitting
train = train %>% select(-row_id)
test = test %>% select(-row_id)

## Initializing score objects
score_object_train = scoreparameters(scoretype = 'bge', train)
score_object_test = scoreparameters(scoretype = 'bge', test)
```

### b)
```{r}
## Infer network on train data
infered_network = orderMCMC(score_object_train)

## Construct graph object with the inferred network
graph_object = graphAM(adjMat = infered_network$DAG, 
                       edgemode='directed', 
                       values=NA)
## Plot graph 
plot(graph_object)

## Evaluate the log BGe score of the test data against the estimated DAG.
DAGscore(score_object_test, infered_network$DAG)
```

### c)
```{r}
## Define function for average BGe score and average number of edges
avg_BGe_score_and_n_edges <- function(am) {
  
  ## Have to load libraries inside the function otherwise it can not find other 
  ## functions even if they are alredy loaded.
  library(BiDAG)
  library(dplyr)
  
  ## Init objects to keep track of BGe score and number of edges
  n_edges = c()
  BGe_scores = c()
  
  for (i in 1:10) {
    ## Split data
    train = data %>% sample_frac(0.8, replace = FALSE) 
    test = anti_join(data, train, by = "row_id")
  
    ## Remove id column used for splitting
    train = train %>% select(-row_id)
    test = test %>% select(-row_id)
    
    ## Initializing score objects
    score_object_train = scoreparameters(scoretype = 'bge', 
                                         train, 
                                         bgepar = list(am = am, aw = NULL))
    score_object_test = scoreparameters(scoretype = 'bge', 
                                        test,
                                        bgepar = list(am = am, aw = NULL))
    
    ## Infer network on train data
    infered_network = orderMCMC(score_object_train)
    
    ## Append number of edges of the inferred network
    n_edges = c(n_edges, sum(infered_network$DAG))
    
    ## Append BGe score
    BGe_scores = c(BGe_scores, DAGscore(score_object_test,
                                        infered_network$DAG))
  }
  
  ## Return mean number of edges and mean BGe score
  return(list("am_value" = am,
              "mean_n_edges" = mean(n_edges), 
              "mean_BGe_score" = mean(BGe_scores)))
}

## Create list of alpha mu values (am values)
am_values = list(10e-5, 10e-3, 10e-1, 10, 10e2)

## Run the each average pair (for each am value) on parallel
## You might have to change this if you computer doesn't have 5 cores!! 
results = mclapply(X = am_values, 
                   FUN = avg_BGe_score_and_n_edges, 
                   mc.cores = 5,  
                   mc.preschedule = FALSE)

## Print results
do.call(rbind, results) 

## The closer x is to 0 the more negative the log(x) is. Therefore the least neg
## average log(BGe_score) corresponds to the largest BGe score (hence am = 1). 

## Plot relearned DAG with whole data set
## Initializing score objects
score_object_data = scoreparameters(scoretype = 'bge', 
                                    data %>% select(-row_id),
                                    bgepar = list(am = 1, aw = NULL))

## Infer network on train data
infered_network = orderMCMC(score_object_data)

## Construct graph object with the inferred network
graph_object = graphAM(adjMat = infered_network$DAG, 
                       edgemode='directed', 
                       values=NA)
## Plot graph 
plot(graph_object)
```


