---
title: "Project 9"
author: "Santiago Castro Dau, June Monge, Rachita Kumar, Sarah Lötscher"
date: '2022-05-05'
output: 
  pdf_document:
    extra_dependencies:
      amsmath
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Problem 23: d-separation 
###  1. Write down all the variables that are d-separated from A given {C,D}
A and G are d-separated since both the paths A-C-G and A-D-G are both blocked 

###  2. Indicate whether each statement is true or false and explain your choice.
#### (a) B is conditionally independent of C given D.
False--> the path B-D-A-C is still active
#### (b) G is conditionally independent of E given D.
False--> the path G-C-A-D-B-F-E is still active
#### (c) C is conditionally independent of F given A.
True--> true since all paths are inactive
#### (d) C is conditionally independent of E given its Markov blanket (of C)
True --> Generally for any node if  conditioned on the markov blanket its going to be independend of the other nodes. Every node is independent of its non-descendants(parents).


## Problem 24: Testing for marginal correlation 
###  Using the data from MVN DAG.rds2, display the observations of A and B in a scatterplot. 
```{r}
library("ggplot2")
library("pcalg")
library("BiDAG")
```

```{r}
#Read in dataset
set.seed(0)
DAG<-readRDS("./MVN_DAG.rds", refhook = NULL)

```

```{r}
#Create the scatterplot

slr.coef<-coef(lm(B~ A,DAG))
qplot(A,B,data=DAG)+theme_classic()+geom_abline(intercept =slr.coef[1], slope = slr.coef[2],colour="blue" )+ ggtitle("Marginal Coorelation")


```
What does the plot suggest about their (marginal) correlation?
There seems to be no trend in the plot which suggests no correlation.
Does it agree with Figure 3?
The results match the Figure since A and B are marginally independent and, thus, their correlation should be 0.

```{r}
#Coorelation Test

cor.test(DAG$A,DAG$B)

```

Use the function cor.test() to test the null hypothesis of no correlation between A and B What is your conclusion?
We can not reject the hypothesis that A and B are marginally uncorrelated as it gives us a p-value of p-value = 0.8401. Therefore, A and B are not correlated.



## Problem 25: Testing for partial correlation 
###Compute the partial correlation ρA,B|C to assess the association between A and B given C as follows:

### 1. Linearly regress A on C (that is, with A as the response variable and C as the explanatory variable). Compute and store the residuals. 
```{r}
regA_C<-lm(A~ C,as.data.frame(DAG))
residualA_C<-residuals(regA_C)
```

### 2. Linearly regress B on C. Compute and store the residuals. 
```{r}
regB_C<-lm(B~ C,as.data.frame(DAG))
residualB_C<-residuals(regB_C)
```

### 3. Plot the residuals of A (regressed on C) against the residuals of B (regressed on C). 
```{r}
slr.coef<-coef(lm(residualB_C~ residualA_C))
qplot(residualA_C,residualB_C)+theme_classic()+geom_abline(intercept =slr.coef[1], slope = slr.coef[2],colour="blue" )+ ggtitle("Partial Correlation ")+labs(y= "B  | C", x = "A | C") 

```
What do you see?
There seems to be a trend in the plot which suggests that  there could be a partial correlation.

### 4. Use the function cor.test() to test the null hypothesis of no correlation between the residualsof A (regressed on C) and the residuals of B (regressed on C)
```{r}
#Coorelation Test

cor.test(residualA_C,residualB_C)

```
What is your conclusion? 
With the p-value of 6.6e-13 we can reject the Null Hypothesis which states that there is no partial correlation. We can say that A and B are not conditionally independent given C.
Does this agree with your expectation based on the underlying DAG in Figure 3?
It agrees with the Figure, as it corresponds to the /explaining away/ scenario in which A and B are parent nodes of C and, thus, given C, they cannot be independent.


## Problem 26: Running the PC algorithm
###Use the function pc() to run the PC algorithm on the data in MVN DAG.rds, and plot the result.
```{r}
sSt <- list(C = cor(DAG),n= nrow(DAG))

PCrun<-pc(suffStat=sSt,indepTest = gaussCItest,labels = colnames(DAG), alpha = 0.01)

plot(PCrun,main="PC Graph")
```

Does the algorithm successfully learn the structure of the data-generating graph in Figure 3? 
Yes the PC algorithm has successfully estimated the correct graph when α =0.01, because it matches both the skeleton and all the colliders. Note that the arrow direction of B & E can not be estimated as it is indistinguishable from that of the true graph: the underlying skeleton is identical for any possible edge orientation that does not form a collider. 
How is the result affected by the significance level α for the conditional independence tests?
The result change with different α values. The higher alpha is, more edges can get accepted so the graph becomes more dense.

## Problem 27: Running the partition MCMC algorithm 
###Run the partition MCMC algorithm using the function partitionMCMC() and plot the result. 
```{r}
param<-scoreparameters(scoretype="bge", DAG)

partmcmc<-partitionMCMC(param)


plot(m2graph(partmcmc$DAG),main="Partition MCMC Algorithm Graph")
```

How is the result affected by the hyper-parameter αμ?
