---
title: "Project 6"
author: "Santiago Castro Dau, June Monge, Rachita Kumar, Sarah Lötscher"
date: '2022-04-11'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Problem 15: Monte Carlo estimation of an expected value
### 1. Show that: $...$



## Problem 16: Sampling in the Rain Network
### a) Derive the expressions for P (C = T | R = T, S = T, W = T), P (C = T | R = F, S =T, W = T), P (R = T | C = T, S = T, W = T) and P (R = T | C = F, S = T, W = T) and compute their values.

\begin{equation*}
\begin{split}
  P(C=T|R=T,S=T,W=T)&=P(C=T|R=T,S=T)&=\frac{P(C=T)P(S=T,R=T|C=T)}{P(S=T,R=T)}\\     &=\frac{P(C=T)P(S=T|C=T)P(R=T|C=T)}{P(C=T)P(S=T|C=T)P(R=T|C=T)+P(C=F)P(S=T|C=F)P(R=T|C=F)}\\
&=\frac{0.5*0.1*0.8}{0.53*0.1*0.8+0.5*0.5*0.2}&=0.4444

P(C=T|R=F,S=T,W=T)&=\frac{P(C=T)P(S=T|C=T)P(R=F|C=T)}{P(C=T)P(S=T|C=T)P(R=F|C=T)+P(C=F)P(S=T|C=F)P(R=F|C=F)}\\
&=\frac{0.5*0.1*0.2}{0.5*0.1*0.2+0.5*0.5*0.8}&=0.04761905



P(R=T|C=T,S=T,W=T)&=\frac{P(R=T|C=T,S=T)P(W=T|R=T,C=T,S=T)}{P(W=T|C=T,S=T)}&=\frac{P(R=T|C=T)P(W=T|R=T,S=T)}{P(W=T|C=T,S=T)}\\  &=\frac{P(R=T|C=T)P(W=T|R=T,S=T)}{P(R=T|C=T)P(W=T|R=T,S=T)+P(R=F|C=T)P(W=T|R=F,S=T)}\\
&=\frac{0.8*0.99}{0.8*0.99+0.2*0.9}&=0.8148

P(R=T|C=F,S=T,W=T)&=\frac{P(R=T|C=F)P(W=T|R=T,S=T)}{P(R=T|C=F)P(W=T|R=T,S=T)+P(R=F|C=F)P(W=T|R=F,S=T)}\\
&=\frac{0.2*0.99}{0.2*0.99+0.8*0.9}&=0.2156863


\end{split}
\end{equation*}

### b) Implement the Gibbs sampler for the Bayesian network 


```{r, echo = FALSE}
library("coda")

```


```{r}

#DOES NOT WORK YET
sampleGibbs <- function(start.a, start.b, n.sims, data){
  # get sum, which is sufficient statistic
  x <- sum(data)
  # get n
  n <- nrow(data)
  # create empty matrix, allocate memory for efficiency
  res <- matrix(NA, nrow = 2, ncol = 2)
  res[1,] <- c(start.a,start.b)
  for (i in 2:n.sims){
    # sample the values
    res[i,1] <- rgamma(1, shape = n+1,
    rate = res[i-1,2]*x+1)
    res[i,2] <- rgamma(1, shape = n+1,
    rate = res[i,1]*x+1)
  }
return(res)
}


# run Gibbs sampler
n.sims <- 100
# return the result (res)
res <- sampleGibbs(.25,.25,n.sims,data)
res
```


### c) Estimate the marginal probability of rain, given that the sprinkler is on and the grass is wet

```{r}

```


### d) Draw 50,000 samples instead of 100 using the Gibbs sampler

```{r}

```

### e) Plot the relative frequencies of R = T and C = T up to each iteration t against t, for two independent runs of the sampler. Suggest a burn-in time based on this plot.

```{r}

```

### f) Apply the Gelman and Rubin test 

```{r}

```

### g) Provide plots for both variables Rain and Cloudy and suggest an interval for drawing approximately independent samples. 

```{r}

```

### h)Re-estimate P (R = T | S = T, W = T) based on samples obtained after the suggested burn-in time and thinning.

```{r}

```

### i)Compute the probability P (R = T | S = T, W = T) analytically. Compare with (c) and (h)
and comment on your results.

```{r}

```


