---
title: "Project 7"
author: "Santiago Castro Dau, June Monge, Rachita Kumar, Sarah Lötscher"
date: '2022-04-28'
output: 
  pdf_document:
    extra_dependencies:
      amsmath
      
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(BiDAG)
library(graph)
```

## Problem 17: Junction Tree Algorithm
### (a) Build the Junction Tree of the network.
#### Moralize graph and trinagulate
```{r}
adj_mat = matrix(1, 
                 ncol = 11, 
                 nrow = 11, 
                 dimnames = list(LETTERS[seq( from = 1, to = 11 )], 
                                 LETTERS[seq( from = 1, to = 11 )]))
adj_mat[lower.tri(adj_mat, diag = TRUE)] = 0

## Changing A
adj_mat[1, ] = c(0,1,1,0,0,0,0,0,0,0,0)
## Changing B
adj_mat[2, ] = c(1,0,1,0,0,0,0,0,0,0,0)
## Changing C
adj_mat[3, ] = c(1,1,0,1,0,1,0,0,0,0,0)
## Changing D
adj_mat[4, ] = c(0,0,1,0,1,1,0,0,0,0,0)
## Changing E
adj_mat[5, ] = c(0,0,0,1,0,1,1,0,0,0,0)
## Changing F
adj_mat[6, ] = c(0,0,1,1,1,0,1,0,1,0,0)
## Changing G
adj_mat[7, ] = c(0,0,0,0,1,1,0,1,0,0,0)
## Changing H
adj_mat[8, ] = c(0,0,0,0,0,0,1,0,0,0,0)
## Changing I
adj_mat[9, ] = c(0,0,0,0,0,1,0,0,0,1,1)
## Changing J
adj_mat[10, ] = c(0,0,0,0,0,0,0,0,1,0,0)
## Changing K
adj_mat[11, ] = c(0,0,0,0,0,0,0,0,1,0,0)

g = graphAM(adjMat = adj_mat, edgemode = 'undirected')
plot(g)
```


#### Build clique grpah

#### Brake cycles

### (b) Write the joint probability P(U) in terms of the cluster and separator potentials.
$$
P(A,B,C,D,E,F,G,H,I,J) = \sum_{A,B,C,D,E,F,G,H,I,J} P(U)
$$

$$
= \sum_{A,B,C,D,E,F,G,H,I,J} \phi(U) =  \frac{\phi(ABC)\phi(DCF)\phi(DEF)\phi(EFG)\phi(GH)\phi(FI)\phi(IS)\phi(IK)}{\phi(C)\phi(DF)\phi(EF)\phi(G)\phi(F)\phi(I)^2}
$$

## Problem 18: Benefit of storing messages
### (a)
The formula for recursively computing the forward and backward messages is:
\begin{align*}
\mu_{\alpha}(x_{n}) &= \sum_{x_{n-1}}\psi_{n-1, n}(x_{n-1}, x_{n})\mu_{\alpha}(x_{n-1})\\
\mu_{\beta}(x_{n}) &= \sum_{x_{n+1}}\psi_{n, n+1}(x_{n}, x_{n+1})\mu_{\beta}(x_{n+1})\\
\end{align*}
Initialization:
\begin{align*}
\mu_{\alpha}(x_{2}) &= \sum_{x_{1}}\psi_{1, 2}(x_{1}, x_{2})\\
\mu_{\beta}(x_{4}) &= \sum_{x_{5}}\psi_{4, 5}(x_{4}, x_{5})\\
%Z &= \sum_{x_{n}}\mu_{\alpha}(x_{n})\mu_{\beta}(x_{n})
\end{align*}

### (b)

The computational complexity for computing the marginal probability 
$P(X_{4} = 1)$ is $O(NK^{2})$. For each node that can assume K values, since $(n - 1)K^{2}$ computations have to be done for the forward message passing and $(N - n)K^{2}$ computations have to be done for the backward message passing, this sums to $(N-1)K^{2}$ computations which is of the order $O(NK^{2})$

### (c)

If the messages are stored, by doing one pass in the forward and one pass in the backward, the total number of steps is reduced to $2(N - 1)K^{2}$, and therefore the order is $O(NK^{2})$. In the general case, the number of steps for $N$ nodes is $N(N-1)K^{2}$ and therefore the order of computational complexity is $O(N^{2} K^{2})$

## Problem 19 (data analysis): Message passing on a chain
### (a) Store clique potentials in an R object
```{r}
clique_potentials = array(dim = c(2, 2, 4), dimnames = list(c("0", "1"), c("0", "1"),
c("Psi12", "Psi23", "Psi34", "Psi45")))
```
### compute clique potentials
```{r}
X1 = c(1/3,2/3)
X2 = c(4/5,2/3)
X3 = c(5/7,1/3)
X4 = c(3/5,2/5)
X5 = c(1/2,7/9)
```

```{r}
clique_potentials = array(dim = c(2, 2, 4), 
                          dimnames = list(c("0", "1"), c("0", "1"),
                                          c("Psi12", "Psi23", "Psi34", "Psi45")))
clique_potentials[1,1,"Psi12"] = (2/3)*(1/5)
clique_potentials[1,2,"Psi12"] = (2/3)*(4/5)
clique_potentials[2,1,"Psi12"] = (1/3)*(1/3)
clique_potentials[2,2,"Psi12"] = (1/3)*(2/3)
clique_potentials[,,"Psi23"] = c(1-X3,X3)
clique_potentials[,,"Psi34"] = c(1-X4,X4)
clique_potentials[,,"Psi45"] = c(1-X5,X5)
```

```{r}
clique_potentials[,,"Psi12"]
```
```{r}
clique_potentials[,,"Psi23"]
```
```{r}
clique_potentials[,,"Psi34"]
```
```{r}
clique_potentials[,,"Psi45"]
```
### (b) Computing forward messages
```{r}
forward_messages = array(dim = c(5,2), dimnames = list(c("X1","X2", "X3","X4","X5"),c("0", "1")))
forward_messages["X1",] = c(1,1)
for(i in 2:5){
  forward_messages[i,]=forward_messages[i-1,]%*%clique_potentials[,,i-1]
}
forward_messages
```

### (d) Computing backward messages
```{r}
backward_messages = array(dim = c(2,5), 
                          dimnames = list(c("0", "1"),
                                          c("X1","X2","X3","X4","X5")))
backward_messages[,"X5"] = c(1,1)
for(i in 4:1){
  backward_messages[,i]=clique_potentials[,,i]%*%backward_messages[,i+1]
}
backward_messages
```
### (e) Compute the marginal probability distribution for each node
```{r}
unnormalised = array(0,dim = c(5,2),
                     dimnames=list(c("p(X1)","p(X2)","p(X3)",
                                     "p(X4)","p(X5)"),
                                   c("0","1")))
for(i in 1:5){
    unnormalised[i,] = forward_messages[i,]*backward_messages[,i]
}
Z = rowSums(unnormalised)
marginal_probability = unnormalised/Z
marginal_probability
```

